Towards Practical First-Order Model Counting

### 2025-01-20 doc/paper/paper.tex
Section 2.4

\todo[inline,caption={}]{
  For the entire paper:
  \begin{itemize}
    \item Finish adding DOIs or URLs to (non-book) references (and reformatting
          indentation) (and check reference formatting) (is the part about
          non-book references correct? Check previous proceedings) (maybe one
          more exception (graded logics)?) (I might be done)
    \item Sentence vs formula: be consistent and not confusing
    \item 15 pages excluding references
    \item Go through the rest of formatting instructions (in GTD)
  \end{itemize}
}
\section{Introduction}

@@

\todo[inline]{Adjust the introduction to the new structure (with more
  subsections)}

@@

\paragraph*{Notation}
We use $\mathbb{N}_{0}$ to represent the set of non-negative integers. In both
algebra and logic, we write $S\sigma$ to denote the application of a
\emph{substitution} $\sigma$ to an expression $S$, where
$\sigma = [x_{1} \mapsto y_{1}, x_{2} \mapsto y_{2}, \dots, x_{n} \mapsto y_{n}]$
signifies the replacement of all instances of $x_{i}$ with $y_{i}$ for all
$i = 1, \dots, n$.

@@

In this section, we will formally define FOMC and its weighted variant. Note
that, although this work focuses on FOMC, for sentences with existential
quantifiers, computing the FOMC using \Cranetwo{} requires the use of WFOMC\@.
For such sentences, Skolemization (described in \cref{sec:crane}) introduces
predicates with non-unary weights that must be accounted for to compute the
correct model count. Additionally, our formulation of the definition of WFOMC is
somewhat complicated by the need to support any number of domains.

@@

\begin{example}[Counting bijections]\label{example:overall}
  Let $\phi$ be the following formula (previously examined by Dilkas and
  Belle~\cite{DBLP:conf/kr/DilkasB23}) that defines predicate $P$ as a bijection
  between two domains $\Gamma$ and $\Delta$:
  \[
    \begin{gathered}
      (\forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y))\land{}\\
      (\forall y \in \Delta\text{. }\exists x \in \Gamma\text{. }P(x, y))\land{}\\
      (\forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z)\land{}\\
      (\forall x, z \in \Gamma\text{. }\forall y \in \Delta\text{. }P(x, y) \land P(z, y) \Rightarrow x = z).
    \end{gathered}
  \]
  Let $\sigma$ be defined as $\sigma(\Gamma) \coloneqq \{\, 1, 2\,\}$, and
  $\sigma(\Delta) \coloneqq \{\,a, b\,\}$. Then $\phi$ has two models:
  \[
    \{\, P(1, a), P(2, b), \neg P(1, b), \neg P(2, a) \,\} \qquad \text{and} \qquad \{\, P(1, b), P(2, a), \neg P(1, a), \neg P(2, b) \,\}.
  \]
\end{example}

@@

\subsection{First-Order Knowledge Compilation}\label{sec:crane}

\todo[inline]{It would be nice to give the necessary introduction to FOKC, or
  \textsc{Crane} (an intuitive explanation early in the paper)}

There are two crucial preprocessing steps that transform an arbitrary sentence
in first-order logic into the form used internally: Skolemization and rewriting
the sentence as a conjunction of clauses. We describe the former in more detail.
\emph{Skolemization}~\cite{DBLP:conf/kr/BroeckMD14} is a procedure that
transforms a formula with existential quantifiers into a formula without
existential quantifiers \emph{with the same WFOMC}. (Note that it is different
from the standard Skolemization that introduces function
symbols~\cite{DBLP:books/daglib/0030198}.)

\begin{example}
  Skolemization transforms
  \begin{equation}\label[formula]{eq:skolemizationinitial}
    \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. } P(x, y)
  \end{equation}
  into
  \begin{equation}\label[formula]{eq:skolemization}
    \begin{aligned}
      (\forall x \in \Gamma\text{. } &Z(x))\land{}\\
      (\forall x \in \Gamma\text{. } \forall y \in \Delta\text{. } &Z(x) \lor \neg P(x, y))\land{}\\
      (\forall x \in \Gamma\text{. } &S(x) \lor Z(x))\land{}\\
      (\forall x \in \Gamma\text{. } \forall y \in \Delta\text{. } &S(x) \lor \neg P(x, y)).
    \end{aligned}
  \end{equation}
  With suitable weights on the new predicates $S/1$ and $Z/1$, the WFOMC of
  \cref{eq:skolemization} is equal to the FOMC of
  \cref{eq:skolemizationinitial}.
\end{example}

@@

Before describing the main contribution of this work, let us review the
essential aspects of FOKC as realised by \textsc{Crane}. The input formula is
compiled into: set $\mathcal{E}$ of equations, map $\mathcal{F}$ from function
names to formulas, and map $\mathcal{D}$ from function names and argument
indices to domains. $\mathcal{E}$ can contain any number of functions, one of
which (denoted by $f$) represents the solution to the FOMC problem. To compute
the FOMC for particular domain sizes, $f$ must be evaluated with those domain
sizes as arguments. $\mathcal{D}$ records this correspondence between function
arguments and domains.

@@

@@

\begin{example}\label{example:overalltwo}
  Let us examine the first solution returned by \Cranetwo{} for the
  bijection-counting problem in \cref{example:overall}. After
  \autoref{line:second}, we have
  \begin{align*}
    \mathcal{E} &= \left\{\,\begin{aligned}f(m, n) &= \sum_{l=0}^{n} \binom{n}{l}{(-1)}^{n-l}g(l, m),\\ g(l, m) &= g(l-1, m) + mg(l-1, m-1)\end{aligned}\,\right\};\\
    \mathcal{D} &= \{\, (f, 1) \mapsto \Gamma, (f, 2) \mapsto \Delta, (g, 1) \mapsto \Delta^{\top}, (g, 2) \mapsto \Gamma \,\},
  \end{align*}
  where $\Delta^{\top}$ is a new domain. (We omit the definition of
  $\mathcal{F}$ as the formulas can get a bit verbose.) Then \FindBaseCases
  identifies two base cases: $g(0, m)$ and $g(l, 0)$. In both cases,
  \CompileWithBaseCases recurses on the formula $\mathcal{F}(g)$ simplified by
  assuming that one of the domains is empty. In the first case, we recurse on
  the formula $\forall x \in \Gamma\text{. }S(x) \lor \neg S(x)$, where $S$ is a
  predicate introduced by Skolemization with weights $w^{+}(S) = 1$ and
  $w^{-}(S) = -1$. Hence, we obtain the base case $g(0, m) = 0^{m}$. In the case
  of $g(l, 0)$, \Propagate{$\psi$, $\Gamma$, $0$} returns an empty formula,
  resulting in $g(l, 0) = 1$.
\end{example}

@@

Finally, the main responsibility of the \Simplify procedure is to handle the
algebraic pattern $\sum_{m=0}^{n}[a \le m \le b] f(m)$. Here: $n$ is a variable,
$a, b \in \mathbb{N}_{0}$ are constants, and $f$ is an expression that may
depend on $m$. Additionally, $[a \le m \le b] = \begin{cases}
  1 & \text{if $a \le m \le b$} \\
  0 & \text{otherwise}
\end{cases}$.
\Simplify transforms this pattern into
$f(a) + f(a+1) + \cdots + f(\min\{\, n, b \,\})$. For instance, in the case of
\cref{example:overalltwo}, \Simplify transforms
$g(l, m) = \sum_{k=0}^{m}[0 \le k \le 1]\binom{m}{k}g(l-1, m-k)$ into
$g(l, m) = g(l-1, m) + mg(l-1, m-1)$.

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item Experiments could be expanded by 1 page (``a more thorough and
          independent practical assessment'')
    \item The difference between \Cranegreedy{} and \Cranebfs{} is in how the
          algorithm chooses which compilation rule to apply to a formula. The
          former uses greedy search: there is a list of rules, and the first
          applicable rule is the one that gets used, disregarding all the
          others. The latter uses a combination of greedy and breadth-first
          search (BFS). That is, each compilation rule is identified as either
          greedy or non-greedy. Greedy rules are applied as soon as possible at
          any stage of the compilation process. BFS is executed over all
          applicable non-greedy rules, identifying the solution that can be
          constructed using the smallest number of such rules.
  \end{itemize}
}

Our empirical evaluation sought to compare the runtime performance of
{\Cranetwo} with the current state of the art, namely \textsc{FastWFOMC} and
\textsc{ForcLift}. It is worth remarking that \textsc{ForcLift} does not support
arbitrary precision, and returns error for cases that requires arbitrary
precision reasoning. Our experiments involve two versions of \Cranetwo{}:
\Cranegreedy{} and \Cranebfs{}. Like its predecessor, \Cranetwo{} has two modes
for applying compilation rules to formulas: one that uses a greedy search
algorithm similar to \textsc{ForcLift} and another that combines greedy and
breadth-first search.

>>>>

\todo[inline,caption={}]{
  For the entire paper:
  \begin{itemize}
    \item Finish adding DOIs or URLs to (non-book) references (and reformatting
          indentation) (and check reference formatting) (is the part about
          non-book references correct? Check previous proceedings) (maybe one
          more exception (graded logics)?) (I might be done)
    \item Sentence vs formula: be consistent and not confusing
    \item 15 pages excluding references
    \item Go through the rest of formatting instructions (in GTD)
    \item Make sure all the references (e.g., to examples are still valid)
    \item Maybe add extra references to subsections
  \end{itemize}
}
\section{Introduction}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item Adjust the introduction to the new structure (with more subsections)
    \item If I need more space elsewhere, shortening the preliminaries to two
          pages could work (I could expand the introduction by 0.5 page)
  \end{itemize}
}

@@

\paragraph*{Notation}
We use $\mathbb{N}_{0}$ to represent the set of non-negative integers. In both
algebra and logic, we write $S\sigma$ to denote the application of a
\emph{substitution} $\sigma$ to an expression $S$, where
$\sigma = [x_{1} \mapsto y_{1}, x_{2} \mapsto y_{2}, \dots, x_{n} \mapsto y_{n}]$
signifies the replacement of all instances of $x_{i}$ with $y_{i}$ for all
$i = 1, \dots, n$. Additionally, for any variable $n$ and
$a, b \in \mathbb{N}_{0}$, let $[a \le n \le b] \coloneqq \begin{cases}
  1 & \text{if $a \le n \le b$} \\
  0 & \text{otherwise}
\end{cases}$.

@@

In this section, we will formally define FOMC and its weighted variant. Note
that, although this work focuses on FOMC, for sentences with existential
quantifiers, computing the FOMC using \Cranetwo{} requires the use of WFOMC\@.
For such sentences, preprocessing (described in \cref{sec:crane}) introduces
predicates with non-unary weights that must be accounted for to compute the
correct model count.

% Additionally, our formulation of the definition of WFOMC is
% somewhat complicated by the need to support any number of domains.

@@

\begin{example}[Counting bijections]\label{example:overall}
  Let us consider the following formula (previously examined by Dilkas and
  Belle~\cite{DBLP:conf/kr/DilkasB23}) that defines predicate $P$ as a bijection
  between two domains $\Gamma$ and $\Delta$:
  \begin{equation}\label[formula]{eq:bijections}
    \begin{gathered}
      (\forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y))\land{}\\
      (\forall y \in \Delta\text{. }\exists x \in \Gamma\text{. }P(x, y))\land{}\\
      (\forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z)\land{}\\
      (\forall x, z \in \Gamma\text{. }\forall y \in \Delta\text{. }P(x, y) \land P(z, y) \Rightarrow x = z).
    \end{gathered}
  \end{equation}
  Let $\sigma$ be defined as $\sigma(\Gamma) \coloneqq \{\, 1, 2\,\}$, and
  $\sigma(\Delta) \coloneqq \{\,a, b\,\}$. Then \cref{eq:bijections} has two
  models:
  \[
    \{\, P(1, a), P(2, b), \neg P(1, b), \neg P(2, a) \,\} \qquad \text{and} \qquad \{\, P(1, b), P(2, a), \neg P(1, a), \neg P(2, b) \,\}.
  \]
\end{example}

@@

\subsection{Crane and First-Order Knowledge Compilation}\label{sec:crane}

\todo[inline]{Make sure the distinction between function definitions and
  equations is clear.}

As our work builds on \textsc{Crane}, in this section we will briefly outline
the steps \textsc{Crane} goes through to compile an \FO{} formula into a set of
function definitions. We divide the inner workings of the algorithm into two
stages: preprocessing and compilation.

\subsubsection{Preprocessing}

The goal of this stage is to transform an arbitrary \FO{} formula into the
format described in \cref{sec:logic}, most importantly by eliminating
existential quantifiers. For example, the first conjunct of
\cref{eq:bijections}, i.e.,
\begin{equation}\label[formula]{eq:skolemizationinitial}
  \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. } P(x, y)
\end{equation}
is transformed into
\begin{equation}\label[formula]{eq:skolemization}
  \begin{aligned}
    (\forall x \in \Gamma\text{. } &Z(x))\land{}\\
    (\forall x \in \Gamma\text{. } \forall y \in \Delta\text{. } &Z(x) \lor \neg P(x, y))\land{}\\
    (\forall x \in \Gamma\text{. } &S(x) \lor Z(x))\land{}\\
    (\forall x \in \Gamma\text{. } \forall y \in \Delta\text{. } &S(x) \lor \neg P(x, y)),
  \end{aligned}
\end{equation}
where $Z/1$ and $S/1$ are two new predicates with $w^{-}(S) = -1$. One can check
that the WFOMC of \cref{eq:skolemizationinitial,eq:skolemization} is the same.

\subsubsection{Compilation}

At this stage, the preprocessed formula is compiled into the set $\mathcal{E}$
of equations and two auxiliary maps $\mathcal{F}$ and $\mathcal{D}$.
$\mathcal{F}$ maps function names to formulas, and $\mathcal{D}$ maps function
names and argument indices to domains. $\mathcal{E}$ can contain any number of
functions, one of which (which we will always denote by $f$) represents the
solution to the FOMC problem. To compute the FOMC for particular domain sizes,
$f$ must be evaluated with those domain sizes as arguments. $\mathcal{D}$
records this correspondence between function arguments and domains.

\begin{example}
  \textsc{Crane} compiles \cref{eq:bijections} for bijection counting into
  \begin{align*}
    \mathcal{E} &= \left\{\,\begin{aligned}f(m, n) &= \sum_{l=0}^{n} \binom{n}{l}{(-1)}^{n-l}g(l, m),\\ g(l, m) &= \sum_{k=0}^{m}[0 \le k \le 1]\binom{m}{k}g(l-1, m-k)\end{aligned}\,\right\};\\
    \mathcal{D} &= \{\, (f, 1) \mapsto \Gamma, (f, 2) \mapsto \Delta, (g, 1) \mapsto \Delta^{\top}, (g, 2) \mapsto \Gamma \,\},
  \end{align*}
  where $\Delta^{\top}$ is a newly introduced domain. (We omit the definition of
  $\mathcal{F}$ as the formulas can get quite verbose.) To compute the number of
  bijections between two sets of cardinality 3, one would evaluate $f(3, 3)$,
  however, the definition of $g$ is incomplete: $g$ is a recursive function
  presented without any base cases. $\mathcal{D}$ encodes that in $f(m, n)$, $m$
  and $n$ represent $|\Gamma|$ and $|\Delta|$, respectively. Similarly, in
  $g(l, m)$, $l$ represents $|\Delta^{\top}|$, and $m$ represents $|\Gamma|$.
\end{example}

Compilation is performed primarily by applying \emph{(compilation) rules} to
formulas. \textsc{Crane} has two modes depending on how the algorithm chooses
which compilation rule to apply to a formula (in case several alternatives are
available). The first option is to use greedy search: there is a list of rules,
and the first applicable rule is the one that gets used, disregarding all the
others. The second option is to use a combination of greedy and
\emph{breadth-first search} (BFS). That is, each compilation rule is identified
as either greedy or non-greedy. Greedy rules are applied as soon as possible at
any stage of the compilation process. BFS is executed over all applicable
non-greedy rules, identifying the solution that can be constructed using the
smallest number of such rules.

@@

@@

\todo[inline]{Connect this with the preliminaries?}

@@

\todo[inline]{The example below builds on \cref{sec:crane}. Shorten it.}

\begin{example}\label{example:overalltwo}
  Let us examine the first solution returned by \Cranetwo{} for the
  bijection-counting problem in \cref{example:overall}. After
  \autoref{line:second}, we have
  \begin{align*}
    \mathcal{E} &= \left\{\,\begin{aligned}f(m, n) &= \sum_{l=0}^{n} \binom{n}{l}{(-1)}^{n-l}g(l, m),\\ g(l, m) &= g(l-1, m) + mg(l-1, m-1)\end{aligned}\,\right\};\\
    \mathcal{D} &= \{\, (f, 1) \mapsto \Gamma, (f, 2) \mapsto \Delta, (g, 1) \mapsto \Delta^{\top}, (g, 2) \mapsto \Gamma \,\},
  \end{align*}
  where $\Delta^{\top}$ is a new domain. (We omit the definition of
  $\mathcal{F}$ as the formulas can get a bit verbose.) Then \FindBaseCases
  identifies two base cases: $g(0, m)$ and $g(l, 0)$. In both cases,
  \CompileWithBaseCases recurses on the formula $\mathcal{F}(g)$ simplified by
  assuming that one of the domains is empty. In the first case, we recurse on
  the formula $\forall x \in \Gamma\text{. }S(x) \lor \neg S(x)$, where $S$ is a
  predicate introduced by preprocessing with weights $w^{+}(S) = 1$ and
  $w^{-}(S) = -1$. Hence, we obtain the base case $g(0, m) = 0^{m}$. In the case
  of $g(l, 0)$, \Propagate{$\psi$, $\Gamma$, $0$} returns an empty formula,
  resulting in $g(l, 0) = 1$.
\end{example}

@@

Finally, the main responsibility of the \Simplify procedure is to handle the
algebraic pattern $\sum_{m=0}^{n}[a \le m \le b] f(m)$. Here: $n$ is a variable,
$a, b \in \mathbb{N}_{0}$ are constants, and $f$ is an expression that may
depend on $m$. \Simplify transforms this pattern into
$f(a) + f(a+1) + \cdots + f(\min\{\, n, b \,\})$. For instance, in the case of
\cref{example:overalltwo}, \Simplify transforms
$g(l, m) = \sum_{k=0}^{m}[0 \le k \le 1]\binom{m}{k}g(l-1, m-k)$ into
$g(l, m) = g(l-1, m) + mg(l-1, m-1)$.

\todo[inline]{Check if the above is still clear given all the changes.}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item Experiments could be expanded by 1 page (``a more thorough and
          independent practical assessment'')
  \end{itemize}
}

Our empirical evaluation sought to compare the runtime performance of
{\Cranetwo} with the current state of the art, namely \textsc{FastWFOMC} and
\textsc{ForcLift}. It is worth remarking that \textsc{ForcLift} does not support
arbitrary precision, and returns error for cases that requires arbitrary
precision reasoning. Our experiments involve two versions of \Cranetwo{}:
\Cranegreedy{} and \Cranebfs{}. Like its predecessor, \Cranetwo{} has two modes
for applying compilation rules to formulas: one that uses a greedy search
algorithm similar to \textsc{ForcLift} and another that combines greedy and
BFS\@.
